---
title: "Statistical power"
author: "Rik Huijzer"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
header-includes:
  - \usepackage{amsmath}
output:   
  bookdown::html_document2:
    css: style.css
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 2
    fig_caption: true
    number_sections: true
    theme: cosmo        
    highlight: kate
    df_print: kable
    code_folding: show
  pdf_document:
    toc: yes
csl: ../style/apa-old-doi-prefix.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE}
library(tidyverse)
library(gridExtra)
```

# Simulations
## Helper functions
```{r}
# Convert array of numbers to dataframe with single column.
numbers2df <- function(X) {
  X <- as.numeric(X) # I don't know why, but plotting libraries need this.
  X <- tibble(X = X) # # ggplot needs a dataframe.
  return(X)
}

# Transparant theme for ggplot which looks better on website.
transparent_theme <- theme(
    panel.background = element_rect(fill = "transparent"), # bg of the panel
    plot.background = element_rect(fill = "transparent", color = NA), # bg of the plot
    panel.grid.major = element_blank(), # get rid of major grid
    panel.grid.minor = element_blank(), # get rid of minor grid
    legend.background = element_rect(fill = "transparent"), # get rid of legend bg
    legend.box.background = element_rect(fill = "transparent"), # get rid of legend panel bg
    legend.key = element_rect(fill = "transparent", colour = NA), # get rid of key legend fill, and of the surrounding
    axis.line = element_line(colour = "black") # adding a black line for x and y axis
)

image.path <- function(filename) sprintf("../images/power/%s.svg", filename)
```

```{r}
# Remove all images to avoid unused images to be kept.
unlink(image.path("*"))
```

## Number of accidents
```{r, message=FALSE, warning=FALSE}
set.seed(2020-02-10)

P_with = 0.0012

run_trial <- function(n_samples, P) {
  X <- runif(1:n_samples, 0, 1)
  # Number of accidents after taking n samples.
  return(sum(X < P))
}

run_trials <- function(n_samples, P, n_trials) {
  results <- 1:n_trials %>% map(function(x) run_trial(n_samples, P))
  return(numbers2df(results))
}

my_round <- function(x) format(round(x, 1), nsmall = 1)
variance <- function() TeX('$\\mu^2')
mean_var <- function(X) sprintf("mean = %s, and variance = %s", my_round(mean(X)), my_round(var(X)))

n_samples = 100000 # sample size per trial.

ggplot_hist <- function(n_samples, name, P, n_trials) {
  df <- run_trials(n_samples, P, n_trials)
  print(mean_var(df$X))
  p <- ggplot(df, aes(X)) + 
    geom_histogram(binwidth = 1) +
    xlab("number of accidents per 100 000 right turns") +
    ylab("frequency") +
    scale_x_continuous(limits = c(0, 200)) + 
    transparent_theme
  ggsave(file=image.path(name), plot=p, width=7, height=4, bg="transparent")
  return(p)
}

ggplot_dens <- function(n_samples, name, P, n_trials) {
  df <- run_trials(n_samples, P, n_trials)
  print(mean_var(df$X))
  p <- ggplot(df, aes(X)) + 
    geom_density() +
    xlab("number of accidents per 100 000 right turns") +
    scale_x_continuous(limits = c(0, 200)) + 
    transparent_theme +
    stat_function(fun=dbinom, args = list(size = 100000, prob = P_with), linetype="dashed") +
    annotate("text", 155, 0.038, label = "Number of accidents") +
    annotate("text", 90, 0.034, label = "B(100 000, 0.0012)")
  ggsave(file=image.path(name), plot=p, width=7, height=4, bg="transparent")
  return(p)
}

ggplot_hist(n_samples, "multiple-years", P_with, 20)
ggplot_dens(n_samples, "many-years", P_with, 100)
ggplot_dens(n_samples, "many-many-years", P_with, 1000)
```

## Comparing binomial and normal distributions
```{r, message=FALSE}
mu.n <- 120

p <- ggplot(data.frame(x = c(0,200)), aes(x)) +
  stat_function(fun=dbinom, args = list(size = 100000, prob = 0.0012), show.legend = TRUE, aes(colour = "B(100 000, 0.0012)")) +
  stat_function(fun=dbinom, args = list(size = 1000, prob = 0.12), show.legend = TRUE, aes(colour = "B(1000, 0.12)")) +
  stat_function(fun=dnorm, args = list(mean = mu.n, sd = sqrt(mu.n)), show.legend = TRUE, aes(colour = "N(120, sqrt(120))")) +
  transparent_theme +
  scale_colour_manual("Distributions", values = c("blue", "black", "green"))
ggsave(file=image.path("binomial-and-normal-distributions"), plot=p, width=7, height=4, bg="transparent")
p
```

## Uniform to normal distribution
```{r}
# Make more generic than above. 
# For the plot take a function.

# Returns a combined plot for a histogram function and distribution function.
plot_hist_and_fn <- function(n, hist_fn, name, dist_fn) {
  df <- numbers2df(1:n %>% map(hist_fn))
  print(mean_var(df$X))
  p = ggplot(df, aes(X)) + 
    geom_histogram(binwidth=1) +
    xlab("number of accidents per 100 000 right turns") +
    ylab("frequency") +
    scale_x_continuous(limits = c(0, 200)) + 
    transparent_theme + 
    stat_function(fun=dist_fn, colour="blue")
  # ggsave(file=image.path(name), plot=p, bg="transparent")
  return(p)
}
```

## Deriving the sample distribution
```{r}
set.seed(2020-02-09) # To allow reproducing this section separately.
```

If we take only a few samples from a normal distribution these will have a higher variance than the normal distribution.

```{r, message=FALSE, warning=FALSE}
sd <- sqrt(mu.n)
generate_sample <- function(n) mean(rnorm(n, mean = mu.n, sd = sd))

generate_samples <- function(n_samples, n_trials) {
  results <- 1:n_trials %>% map(function(x) generate_sample(n_samples))
  return(results)
}

ggplot_dens <- function(n_samples, n_trials, name) {
  df <- numbers2df(generate_samples(n_samples, n_trials))
  print(mean_var(df$X))
  p <- ggplot(df, aes(X)) + 
    geom_histogram(binwidth=1) +
    xlab("number of accidents") +
    scale_x_continuous(limits = c(0, 200)) + 
    transparent_theme
  ggsave(file=image.path(name), plot=p, width=7, height=4, bg="transparent")
  return(p)
}

n_trials <- 10000
ggplot_dens(1, n_trials, "one-sample-distribution")
ggplot_dens(2, n_trials, "two-sample-distribution")

dnorm(mu.n, mean=mu.n, sd=sd)
dnorm(120, mean=mu.n, sd=sd)

# Remove population mean
generate_Si <- function(n) rnorm(n, mean=mu.n, sd=sd)
df <- tibble(Samples = 1:1e5 %>% map(function(x) generate_Si(2)))
df$Mean <- as.numeric(df$Samples %>% map(mean))
df$Sd <- as.numeric(df$Samples %>% map(function(x) sd(x, na.rm=TRUE)))
df$X <- as.numeric(df$Mean - mu.n)

p <- ggplot(df, aes(X)) + 
  geom_histogram(binwidth = 0.1) +
  xlab("number of accidents") +
  transparent_theme
ggsave(file=image.path("two-sample-without-mean"), plot=p, width=7, height=4, bg="transparent")
p

# Remove distribution's standard deviation
df$W <- df$X / (sqrt(65) / sqrt(2))
p <- ggplot(df, aes(W)) + 
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(limits = c(-4, 4)) + 
  xlab("number of accidents") +
  transparent_theme
ggsave(file=image.path("two-sample-without-sd"), plot=p, bg="transparent")
p

# Add t distribution
p <- ggplot(df, aes(W)) + 
  geom_density() +
  xlab("number of accidents") +
  transparent_theme + 
  scale_y_continuous(limits = c(0, 0.42)) + 
  scale_x_continuous(limits = c(-4, 4)) + 
  stat_function(fun=dt, args=list(df = 1), show.legend = TRUE, aes(colour = "T(1)"))
ggsave(file=image.path("two-sample-with-t-1"), plot=p, width=7, height=4, bg="transparent")
p
```
**Still incorrect, don't know why**

## Comparing normal and t-distributions
```{r, message=FALSE}
p <- ggplot(data.frame(x = c(-4,4)), aes(x)) +
  stat_function(fun=dnorm, args = list(mean = 0, sd = 1)) +
  stat_function(fun=dt, args = list(df = 5), linetype="dotted") +
  stat_function(fun=dt, args = list(df = 1), linetype="dashed") +
  transparent_theme +
  annotate("text", 0, 0.42, label = "N(0, 1)") +
  annotate("text", 0, 0.35, label = "T(5)") +
  annotate("text", 0, 0.28, label = "T(1)") +
  scale_colour_manual("Distributions", values = c("blue", "black", "green", "red"))
ggsave(file=image.path("normal-and-t-distributions"), width=7, height=4, plot=p, bg="transparent")
p
```

## Some samples
```{r}
set.seed(2020-02-13)
n_samples <- 12
samples <- tibble(
  index = 1:n_samples,
  without = as.numeric(map(rnorm(n_samples, mean = 100, sd = sqrt(mu.n)), my_round)),
  with = as.numeric(map(rnorm(n_samples, mean = mu.n, sd = sqrt(mu.n)), my_round))
)
head(samples, 10)
```

## Single calculation example
```{r}
n = 2
x = samples$with[1:n]
t = (mean(x) - 100) /(sd(x) / sqrt(n))
t
s <- sd(x)
sprintf("s: %s", s)

# Two-sided 95% CI
qt(0.975, df=n-1)

t.test(x, mu = 100, alternative="two.sided", var.equal = TRUE)
```
## t-distribution one versus two sided
```{r}
n = 2
right.one.sided <- qt(0.80, df=n-1)
right.one.sided

p <- ggplot(data.frame(x = c(-4,4)), aes(x)) +
  stat_function(fun=dt, args = list(df = n-1), colour="black") +
  stat_function(fun=dt, args = list(df = n-1), xlim=c(-4,right.one.sided), geom="area", alpha=0.4, fill="blue", show.legend = FALSE) +
  stat_function(fun=dt, args = list(df = n-1), xlim=c(-right.one.sided,right.one.sided), geom="area", alpha=0.2, fill="blue", show.legend = TRUE) +
  transparent_theme +
  annotate("text", -2.8, 0.38, label = "20%") +
  annotate("text", -right.one.sided - 0.25, 0.35, label = "-a") +
  geom_vline(aes(xintercept=-right.one.sided), colour="blue", alpha=0.2, linetype="dashed") +
  annotate("text", 0, 0.38, label = "60%") +
  geom_vline(aes(xintercept=right.one.sided), colour="blue", alpha=0.2, linetype="dashed") +
  annotate("text", 2.8, 0.38, label = "20%") +
  ylab("probability") +
  annotate("text", right.one.sided - 0.25, 0.35, label = "a") +
  theme(legend.position = "none")
ggsave(file=image.path("t-one-versus-two-sided"), width=8, height=4, plot=p, bg="transparent")
p
```

## Output for two samples plotted
```{r, message=FALSE}
one.sided.95 <- qt(0.95, df=1)
two.sided.95 <- qt(0.975, df=1)
round.3 <- function(x) format(round(x, 3), nsmall=3)

p <- ggplot(data.frame(x = c(-0.5,5)), aes(x)) +
  stat_function(fun=dt, args = list(df = 1), show.legend = FALSE, aes(colour = "T(n-1) = T(1)")) +
  transparent_theme +
  geom_vline(aes(xintercept=one.sided.95, colour = sprintf("x = %s (one-sided 95%% critical value)", round.3(one.sided.95))), linetype="dashed") +
  geom_vline(aes(xintercept=two.sided.95, colour = sprintf("x = %s (two-sided 95%% critical value)", round.3(two.sided.95))), linetype="dashed") +
  geom_vline(aes(xintercept=t, colour = sprintf("x = %s (calculated t-value)", round.3(t))), linetype="dashed") +
  ylab("probability") + 
  theme(legend.title = element_blank())
ggsave(file=image.path("t-and-critical-values"), width=7, height=3, plot=p, bg="transparent")
p
```

## Doing multiple t-tests
```{r}
get.p <- function(n, alternative, use.population=FALSE) {
  X <- samples$without[1:n]
  Y <- samples$with[1:n]
  if (use.population) {
    return(t.test(Y, mu = 100, alternative = alternative, var.equal = TRUE)$p.value)
  } else {
    return(t.test(Y, X, alternative = alternative, var.equal = TRUE)$p.value)
  }
}

plot.power <- function(name, use.population=FALSE) {
  df <- tibble(x = 2:n_samples)
  df$p.greater <- as.numeric(df$x %>% map(function(x) get.p(x, "greater", use.population = use.population)))
  df$p.two.sided <- as.numeric(df$x %>% map(function(x) get.p(x, "two.sided", use.population = use.population)))
  
  df1 <- tibble(x = 2:n_samples)
  df1$p <- as.numeric(df$x %>% map(function(x) get.p(x, "two.sided", use.population = use.population)))
  df1$alternative = "two.sided"
  
  df2 <- tibble(x = 2:n_samples)
  df2$p <- as.numeric(df$x %>% map(function(x) get.p(x, "greater", use.population = use.population)))
  df2$alternative = "greater"
  
  df3 <- tibble(x = 2:n_samples)
  df3$p <- as.numeric(df$x %>% map(function(x) get.p(x, "less", use.population = use.population)))
  df3$alternative = "less"
  
  df <- rbind(df1, df2, df3)
  
  # To compare manual calculations to.
  head(df2, 3)
  head(df1, 3)
  
  p <- ggplot(df, aes(x, p)) +
    geom_point(aes(colour = alternative)) +
    scale_x_continuous(breaks = 1:n_samples) +
    geom_hline(yintercept = 0.05) + 
    geom_line(aes(colour = alternative)) +
    annotate("text", n_samples - 1, 0.05, vjust = -1, label = "p = 0.05") +
    xlab("sample size") +
    ylab("p-value") + 
    transparent_theme 
  ggsave(file=image.path(name), plot=p, bg="transparent")
  return(p)
}

plot.power("power-one-sample", use.population = TRUE)
plot.power("power-two-sample", use.population = FALSE)
```

## Manual cohen's d calculation
```{r, message=FALSE}
library(pwr)
# Checkout the generated html to see variable name and output.
samples_a <- samples$with[1:2]
samples_a
samples_b <- samples$without[1:2]
samples_b
s_a <- sd(samples_a)
s_a
s_b <- sd(samples_b)
s_b
m_a <- mean(samples_a)
m_a
m_b <- mean(samples_b)
m_b

s_p <- sqrt((s_a^2 + s_b^2) / 2)
s_p
d <- (m_a - m_b) / s_p
d
pwr.t.test(n=2, d=d, sig.level=0.05, alternative="two.sided")
  
```

## Plotting the value for d over multiple samples
```{r}
plot.d <- function(n, name) {
  
  get.d <- function(n) {
    S_a <- samples$with[1:n]
    S_b <- samples$without[1:n]
    
    s_p <- function(S_a, S_b) sqrt((sd(S_a)^2 + sd(S_b)^2) / 2)
    d <- function(S_a, S_b) (mean(S_a) - mean(S_b)) / s_p(S_a, S_b)
    
    print(sprintf("n: %s, s_p: %s, d: %s", n, s_p(S_a, S_b), d(S_a, S_b)))
    return(d(S_a, S_b))
  }
  
  get.d(2)
  df <- tibble(x = 2:n)
  df$d <- as.numeric(df$x %>% map(get.d))
  
  print(head(df, 4))
  
  p <- ggplot(df, aes(x, d)) +
    geom_point() +
    geom_line() +
    scale_x_continuous(breaks = 1:n_samples) +
    xlab("sample size") +
    expand_limits(y=0) + 
    ylab("Cohen's d") + 
    transparent_theme 
  ggsave(file=image.path(name), width=7, height=3, plot=p, bg="transparent")
  return(p)
}

plot.d(12, "d")
```

## Power analysis (as simple and straightforward as possible)
```{r}
# Estimating effect size as a function of n, d, and alpha
# This assumes that our effect size is 1.
# Combine plot with one for effect size is 0.5.
pwr.t.test(n=2, d=1, sig.level=0.05, alternative="two.sided")

partial.test.8 <- function(n) pwr.t.test(n=n, d=0.8, sig.level=0.05, alternative="two.sided")$power
partial.test.5 <- function(n) pwr.t.test(n=n, d=0.5, sig.level=0.05, alternative="two.sided")$power
partial.test.2 <- function(n) pwr.t.test(n=n, d=0.2, sig.level=0.05, alternative="two.sided")$power

p <- ggplot(data.frame(x = c(2, 200)), aes(x)) +
  stat_function(fun=partial.test.8) +
  stat_function(fun=partial.test.5) +
  stat_function(fun=partial.test.2) +
  ylab("power") +
  xlab("sample size") +
  transparent_theme +
  annotate("text", 25, 0.95, label="d = 0.80") +
  annotate("text", 60, 0.65, label="d = 0.50") + 
  annotate("text", 120, 0.4, label="d = 0.20")
ggsave(file=image.path("varying-d"), width=7, height=4, plot=p, bg="transparent")
p
```

```{r}
# For example in text
pwr.t.test(d=0.5, power=0.6, sig.level=0.05, alternative="two.sided")
pwr.t.test(n=2, d=0.5, sig.level=0.05, alternative="two.sided")
```

```{r}
# Type I error versus type II error
partial.test.n2 <- function(a) pwr.t.test(n=2, d=0.5, sig.level=a, alternative="two.sided")$power
partial.test.n20 <- function(a) pwr.t.test(n=20, d=0.5, sig.level=a, alternative="two.sided")$power
partial.test.n80 <- function(a) pwr.t.test(n=80, d=0.5, sig.level=a, alternative="two.sided")$power

p <- ggplot(data.frame(x = c(0,1)), aes(x)) +
  stat_function(fun=partial.test.n2) +
  stat_function(fun=partial.test.n20) +
  stat_function(fun=partial.test.n80) +
  ylab("power") +
  xlab("significance level") +
  scale_x_continuous(trans = "reverse") +
  transparent_theme +
  annotate("text", 0.50, 0.45, label="n = 2") +
  annotate("text", 0.32, 0.65, label="n = 20") +
  annotate("text", 0.12, 0.85, label="n = 80") +
  geom_vline(aes(xintercept=0.05), linetype="dashed") +
  annotate("text", 0.13, 0.02, label="p = 0.05")
ggsave(file=image.path("varying-n"), width=7, height=4, plot=p, bg="transparent")
p

```

```{r}
pwr.t.test(n=2, d=0.5, sig.level=0.05, alternative="two.sided")$power
pwr.t.test(n=20, d=0.5, sig.level=0.05, alternative="two.sided")$power
pwr.t.test(n=80, d=0.5, sig.level=0.05, alternative="two.sided")$power
```

# Appendix

## R version
```{r version}
sessionInfo()
```